### 2014-12-16 text/Chapter02.tex
Spell checking

Another pathway to coeffects leads through the semantics of effectful and context-dependent 
computations. In a pioneering work, Moggi \cite{monad-notions} showed that effects (including
partiality, exceptions, non-determinism and I/O) can be modelled uisng the category theoretic
notion of \emph{monad}.

@@

In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
of the (unevaluated) expression. We write $\square^{\cclrd{\Psi}} \tau$ for a type of expressions
that requires a context $\Psi$. The type is a comonadic counterpart to \emph{indexed monads} used by 
Wadler and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language 
that tracks context-dependence of computations in a type system.

@@

The first pathway follows as a dualization of well-known effect systems. However, this is not simply 
a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda 
abstraction differently. The second pathway follows by extending comonadic semantics of context-dependent 
computations with indexing and building a type system analogous to effect system from the ``marriage of 
effects and monads''. Finally, the third pathway starts with substructural type systems. Coeffect
systems naturally arise by annotating bunches in bunched logics with additional information. In this
thesis, we mostly follow the first two approaches.

>>>>

Another pathway to coeffects leads through the semantics of effectful and context-dependent 
computations. In a pioneering work, Moggi \cite{monad-notions} showed that effects (including
partiality, exceptions, non-determinism and I/O) can be modelled using the category theoretic
notion of \emph{monad}.

@@

In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variables
of the (unevaluated) expression. We write $\square^{\cclrd{\Psi}} \tau$ for a type of expressions
that requires a context $\Psi$. The type is a comonadic counterpart to \emph{indexed monads} used by 
Wadler and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language 
that tracks context-dependence of computations in a type system.

@@

The first pathway follows as a dualisation of well-known effect systems. However, this is not simply 
a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda 
abstraction differently. The second pathway follows by extending comonadic semantics of context-dependent 
computations with indexing and building a type system analogous to effect system from the ``marriage of 
effects and monads''. Finally, the third pathway starts with substructural type systems. Coeffect
systems naturally arise by annotating bunches in bunched logics with additional information. In this
thesis, we mostly follow the first two approaches.

### 2014-12-16 text/Chapter03.tex
Spell checking

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and constrasts it with a better known notion of 
  \emph{effect}. It explains what is the nature of properties that can be tracked using 
  coeffect systems presented in this theis.

@@

\item In addition, the coeffect systems for tracking the number of accessed past values in 
  data-flow languagges (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
  presents novel results and can be used to optimize data-flow programs.
\end{itemize}

@@

When introducing coeffect systems in Section~\ref{sec:path-effects-coeff}, we related coeffect systems
with effect systems. Effect systems track how a program affects the environment, or, in other words 
capture some \emph{output impurity}. In contrast, coeffect systems track what a program requires from 
the execution environment, or \emph{input impurity}.

@@

The $\lambda$-calculus is asymmetric. It maps a context with \emph{multiple} variables to a 
\emph{single} result. An expression with $n$ free variables of types $\tau_i$ can be modelled by a function 
$\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
the function arrow. However, in the underlyng categorical model, effects are attached to the result
$\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.

@@

We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.

@@

When using the usualmonadic semantics based on the reader monad, the semantics of the (\emph{abs})
rule would be modified as follows:
%
\begin{equation*}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = 
      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)      
\end{array}
\end{equation*}
%
Note that the declaration site dictionary
is ignored and the body is called with only the dictionary provided by the call site. This is
a consequence of the fact that monadic functions are always pure values created using monadic
\emph{unit}, which turns a function $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$ into a monadic
computation with no side-effects $\mtyp{\cclrd{\emptyset}}{\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}}$.

@@

@@

The systems work in the same way as the examples discussed already. For example, consider the
tainting example with the \ident{query} function calling an SQL database. To capture such 
tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
$\cclrd{\ident{U}}$ for \emph{untainted}. Accessing a variable marks it as untained,
but using an expression that depends on some variable in certain dangerous contexts -- such 
as in arguments of \ident{query} -- does introduce a taint on all the variables contributing to
the expression. This is captured using the standard application rule (\emph{app}):
%
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma}{\aclrd{r}} \vdash \ident{query} : \ident{string} \xrightarrow{\cclrd{\ident{T}}} \ident{Table} \qquad
    \coctx{\ident{id}:\ident{string}}{ \alift{\cclrd{\ident{U}}}} \vdash \ident{id} : \ident{string} }
  { \coctx{\Gamma, \ident{id}:\ident{string}}{ \aclrd{r} \cons \alift {\cclrd{\ident{T}}} \vdash \ident{query}(\str{...}, \ident{id}) : \ident{Table} } }
\end{equation*}
%
The derivation assumes that \ident{query} is a standard function that requires the parameters
to be tainted (it does not have to be a built-in language construct). The argument is a 
variable and so it is not tainted in the assumptions.

@@

Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

@@

This chapter served two purposes. The first aim was to present existing work on programming 
languages and systems that include some notion of \emph{context}. Because there was no well-known
abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses 
as well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 
security properties.

>>>>

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and contrasts it with a better known notion of 
  \emph{effect}. It explains what is the nature of properties that can be tracked using 
  coeffect systems presented in this thesis.

@@

\item In addition, the coeffect systems for tracking the number of accessed past values in 
  data-flow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
  presents novel results and can be used to optimize data-flow programs.
\end{itemize}

@@

When introducing coeffect systems in Section~\ref{sec:path-effects-coeff}, we related coeffect systems
with effect systems. Effect systems track how a program affects the environment, or, in other words 
capture some \emph{output impurity}. In contrast, coeffect systems track what a program requires from 
the execution envionment, or \emph{input impurity}.

@@

The $\lambda$-calculus is asymmetric. It maps a context with \emph{multiple} variables to a 
\emph{single} result. An expression with $n$ free variables of types $\tau_i$ can be modelled by a function 
$\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
the function arrow. However, in the underlying categorical model, effects are attached to the result
$\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.

@@

We note thatt an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.

@@

When using the usual monadic semantics based on the reader monad, the semantics of the (\emph{abs})
rule would be modified as follows:
%
\begin{equation*}
\hspace{-0.5em}
\begin{array}{l}
  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = 
      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)      
\end{array}
\end{equation*}
%
Note that the declaration site dictionary
is ignored and the body is called with only the dictionary provided by the call site. This is
a consequence of the fact that monadic functions are always pure values created using monadic
\emph{unit}, which turns a function $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$ into a monadic
computation with no side-effects $\mtyp{\cclrd{\emptyset}}{\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}}$.

@@

%----------------------------------------------------------------------------------------------------

@@

The systems work in the same way as the examples discussed already. For example, consider the
tainting example with the \ident{query} function calling an SQL database. To capture such 
tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
$\cclrd{\ident{U}}$ for \emph{untainted}. Accessing a variable marks it as untainted,
but using an expression that depends on some variable in certain dangerous contexts -- such 
as in arguments of \ident{query} -- does introduce a taint on all the variables contributing to
the expression. This is captured using the standard application rule (\emph{app}):
%
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma}{\aclrd{r}} \vdash \ident{query} : \ident{string} \xrightarrow{\cclrd{\ident{T}}} \ident{Table} \qquad
    \coctx{\ident{id}:\ident{string}}{ \alift{\cclrd{\ident{U}}}} \vdash \ident{id} : \ident{string} }
  { \coctx{\Gamma, \ident{id}:\ident{string}}{ \aclrd{r} \cons \alift {\cclrd{\ident{T}}} \vdash \ident{query}(\str{...}, \ident{id}) : \ident{Table} } }
\end{equation*}
%
The derivation assumes that \ident{query} is a standard function that requires the parameters
to be tainted (it does not have to be a built-in language construct). The argument is a 
variable and so it is not tainted in the assumptions.

@@

Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
a sound system with region-based memory management for low-level code that can be easily 
compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
who developed an effect system (as discussed in Section~\ref{sec:paath-sem-effects}) that uses
lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

@@

This chapter served two purposes. The first aim was to present existing work on programming 
languages and systems that include some notion of \emph{context}. Because there was no well-known
abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses 
as well as approaches based on monads. We looked at a number of applications including Haskell's implicit 
parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 
security properties.
